---
title: "US Public Transit System Data Analysis"
author: "Cindy Li"
---

## Introduction

When I was in Japan, I got to ride some of the most efficient transit systems in the world, characterized by its punctuality, its service, and the large numbers of passengers. Here in the United States, public transit is a whole different story. At the end of this analysis, I want to figure out what the most efficient transit system in the country is. How do I define efficiency? To me, efficiency can be broken down into ridership (trips, miles, the functionality) and financial efficiency (revenue, expenses).

## Data Sets

Let's dive into it by exploring the following data sets:

-   2022 Fare Revenue

-   2022 Expenses

-   Ridership

I will be extracting data on fares, expenses,

### Fare Revenue Data

```{r echo=FALSE, message=FALSE}
library(tidyverse)
FARES <- readxl::read_xlsx("2022_fare_revenue.xlsx") |>
  select(-`State/Parent NTD ID`, 
         -`Reporter Type`,
         -`Reporting Module`,
         -`TOS`,
         -`Passenger Paid Fares`,
         -`Organization Paid Fares`) |>
  filter(`Expense Type` == "Funds Earned During Period") |>
  select(-`Expense Type`)
library(DT)
sample_n(FARES, 1000) |>
  DT::datatable()
```

### Expenses Data

```{r echo=FALSE, message=FALSE}
EXPENSES <- readr::read_csv("2022_expenses.csv") |>
  select(`NTD ID`, 
         `Agency`,
         `Total`, 
         `Mode`) |>
  mutate(`NTD ID` = as.integer(`NTD ID`)) |>
  rename(Expenses = Total) |>
  group_by(`NTD ID`, `Mode`) |>
  summarize(Expenses = sum(Expenses)) |>
  ungroup()
library(DT)
sample_n(EXPENSES, 1000) |>
  DT::datatable()
```

### Financials Data

We can inner join the Revenue and Expenses data into a more comprehensiive financials data set that we can do our analysis with.

```{r echo=FALSE, message=FALSE}
FINANCIALS <- inner_join(FARES, EXPENSES, join_by(`NTD ID`, `Mode`))
```

Here's a sample of the Financials Data.

```{r echo=FALSE, message=FALSE}
library(DT)
sample_n(FINANCIALS, 1000) |>
  DT::datatable()
```

### Trips Data

From the ridership data, I will be extracting information on public transportation trips taken by unlinked passengers.

```{r echo=FALSE, message=FALSE}
TRIPS <- readxl::read_xlsx("ridership.xlsx", sheet="UPT") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="UPT") |>
  drop_na() |>
  mutate(month=my(month))
library(DT)
sample_n(TRIPS, 1000) |>
  DT::datatable()
```

### Miles Data

Also from the ridership data, I will be extracting information on the vehicle revenue miles.

```{r echo=FALSE, message=FALSE}
MILES <- readxl::read_xlsx("ridership.xlsx", sheet="VRM") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="VRM") |>
  drop_na() |>
  group_by(`NTD ID`, `Agency`, `UZA Name`, 
           `Mode`, `3 Mode`, month) |>
  summarize(VRM = sum(VRM)) |>
  ungroup() |>
  mutate(month=my(month))
library(DT)
sample_n(MILES, 1000) |>
  DT::datatable()
```

### Usage Data

We can inner join the Trips and Miles data together using NTD ID into a data set on the usage.

```{r echo=FALSE, message=FALSE}
USAGE <- inner_join(TRIPS, MILES) |>
  mutate(`NTD ID` = as.integer(`NTD ID`))
```

[**Attributes**]{.underline}

Some of the attribute namings are unclear so let's use the [FTA Glossary](https://www.transit.dot.gov/ntd/national-transit-database-ntd-glossary) to interpret the data.

**Renaming**

-   renaming UZA Name to metro_area

-   replacing the modes with their respective full names

-   renaming UPT to unlinked_passenger_trips

-   renaming VRM to vehicle_revenue_miles

```{r echo=FALSE, message=FALSE}
USAGE <- USAGE |>
  rename(metro_area = `UZA Name`) |>
  mutate(Mode=case_when(
    Mode == "HR" ~ "Heavy Rail",
    Mode == "YR" ~ "Heavy Rail",
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferryboat",
    Mode == "MB" ~ "Bus",
    Mode == "SR"~ "Streetcar",
    Mode == "TB" ~ "Trolleybus",
    Mode == "VP"~ "Vanpool",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "LR" ~ "Light Rail",
    Mode == "YR" ~ "Hybrid Rail",
    Mode == "MG" ~ "Monorail",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "TR" ~ "Aerial Tramways",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "PB" ~ "Publico",
    Mode == "CC" ~ "Cable Car",
    TRUE ~ "Unknown"))
USAGE <- USAGE |>
  rename(unlinked_passenger_trips = `UPT`) |>
  rename(vehicle_revenue_miles = `VRM`)
```

**VRM** (vehicle revenue miles): The miles that vehicles travel while in revenue service.

**UPT** (unlinked passenger trips): The number of passengers who board public transportation vehicles. Passengers are counted each time they board vehicles no matter how many vehicles they use to travel from their origin to their destination.

Now,, here's a sample of the processed Usage Data.

```{r echo=FALSE, message=FALSE}
library(DT)
sample_n(USAGE, 1000) |>
  DT::datatable()
```

### Join Usage and Financial Data

Next, we can join the Usage and Financial Data sets.

In order to do so, we need to get the Usage for 2022 in order to match the 2022 Financial data.

```{r echo=FALSE, message=FALSE}
USAGE_2022_ANNUAL <- USAGE|> 
  mutate(year = year(month)) |> 
  filter(year == 2022) |> 
  group_by(`NTD ID`, Agency, metro_area, Mode) |> 
  summarize( UPT = sum(unlinked_passenger_trips), VRM = sum(vehicle_revenue_miles)) |> 
  ungroup()
```

Since we are joining on Mode, we need to convert the modes of the Financials data as well.

```{r echo=FALSE, message=FALSE}
FINANCIALS <- FINANCIALS |> 
  mutate(Mode=case_when( Mode == "HR" ~ "Heavy Rail", Mode == "YR" ~ "Heavy Rail", Mode == "DR" ~ "Demand Response", Mode == "FB" ~ "Ferryboat", Mode == "MB" ~ "Bus", Mode == "SR"~ "Streetcar", Mode == "TB" ~ "Trolleybus", Mode == "VP"~ "Vanpool", Mode == "CB" ~ "Commuter Bus", Mode == "RB" ~ "Bus Rapid Transit", Mode == "LR" ~ "Light Rail", Mode == "YR" ~ "Hybrid Rail", Mode == "MG" ~ "Monorail", Mode == "CR" ~ "Commuter Rail", Mode == "AR" ~ "Alaska Railroad", Mode == "TR" ~ "Aerial Tramways", Mode == "IP" ~ "Inclined Plane", Mode == "PB" ~ "Publico", Mode == "CC" ~ "Cable Car", TRUE ~ "Unknown"))
```

After that, we can LEFT JOIN the two data sets.

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS <- left_join(USAGE_2022_ANNUAL, FINANCIALS, join_by(`NTD ID`, Mode)) |> 
  drop_na()
```

Let's take a look at the Usage and Financials data.

```{r echo=FALSE, message=FALSE}
library(DT)
sample_n(USAGE_AND_FINANCIALS, 1000) |>
  DT::datatable()
```

## Project Outcomes

I used summary statistics to explore the data sets processed above to extract insights that can shed light on efficiency of the US Public Transit Systems.

Libraries: tidyverse, dplyr

Let's see what the data can tell us about public transit in the US looking at transit Usage and Financial data.

### Vehicle Revenue Miles

**What transit agency had the most total VRM in this sample?**

*MTA New York City Transit with 10832855350 total miles*

```{r echo=FALSE, message=FALSE}
library(dplyr)
USAGE |> 
  group_by(Agency) |>
  summarise(total_VRM = sum(vehicle_revenue_miles)) |>
  arrange(desc(total_VRM)) |>
  slice(1)
```

**What transit mode had the most total VRM in this sample?**

*The Bus at 49444494088 total miles*

```{r echo=FALSE, message=FALSE}
USAGE |>
  group_by(Mode) |>
  summarise(total_VRM = sum(vehicle_revenue_miles)) |>
  arrange(desc(total_VRM)) |>
  slice(1)
```

**What mode of transport had the longest average trip in May 2024?**

*The Heavy Rail did with 2654864 average miles*

```{r echo=FALSE, message=FALSE}
USAGE |> 
  filter(month == "2024-05-01") |> 
  group_by(Mode) |> 
  summarise(average_VRM = mean(vehicle_revenue_miles)) |> 
  arrange(desc(average_VRM)) |>
  slice(1)
```

### Unlinked Passenger Trips

**How many trips were taken on the NYC Subway (Heavy Rail) in May 2024?**

*A total of 237383777 trips were taken.*

```{r echo=FALSE, message=FALSE}
TRIPS |> 
  filter(Mode == "HR", month == "2024-05-01") |> 
  summarise(total_trips = sum(UPT))
```

**How much did NYC subway ridership fall between April 2019 and April 2020?**

*Ridership fell by 296864650 between April 2018 and April 2020.*

```{r echo=FALSE, message=FALSE}
april_2020 <- USAGE |> 
  filter(Mode == "Heavy Rail") |> 
  filter(month == "2020-04-01") |> 
  summarise(total_riders = sum(unlinked_passenger_trips))

april_2019 <- USAGE |> 
  filter(Mode == "Heavy Rail") |> 
  filter(month == "2019-04-01") |> 
  summarise(total_riders = sum(unlinked_passenger_trips))

difference = abs(april_2020 - april_2019)
print(difference)
```

**Which transit system (agency and mode) had the most UPT in 2022?**

*The MTA New York City Transit Heavy Rail had the most UPT in 2022 at 1793073801.*

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS |> select(Agency, Mode, UPT) |> 
  arrange(desc(UPT))|>
  slice(1)
```

### Three more interesting transit facts

**Which month had the highest number of average trips between 2002 and 2024?**

*October with an average of 768205 trips.*

```{r echo=FALSE, message=FALSE}
USAGE <- USAGE |> mutate(month_number = month(month))
USAGE |> group_by(month_number) |> summarise(avg_UPT = mean(unlinked_passenger_trips)) |> arrange(desc(avg_UPT)) |>
  slice(1)
USAGE <- USAGE |> select(-month_number)
```

**Which metro area had the most unlinked passenger trips?**

*The New York, Jersey City, and Newark area has the greatest total UPT.*

```{r echo=FALSE, message=FALSE}
USAGE |> group_by(metro_area) |> 
  summarise(total_UPT = sum(unlinked_passenger_trips)) |> 
  arrange(desc(total_UPT)) |> 
  slice(1)
```

**Which metro area offers the most modes of transit?**

*San Francisco--Oakland, CA with 13 Modes*

```{r echo=FALSE, message=FALSE}
USAGE |> group_by(metro_area) |> 
  summarise(mode_count = n_distinct(Mode)) |> 
  arrange(desc(mode_count)) |> 
  slice(1)
```

```{r echo=FALSE, message=FALSE}
distinct(USAGE |> select(Mode, metro_area) |> filter(metro_area == "San Francisco--Oakland, CA"))
```

## Financial Efficiency

### Farebox Recovery Among Major Systems

Farebox recovery is defined as the highest ratio of Total Fares to Expenses and can be used to measure efficiency.

**Which transit system (agency and mode) had the highest farebox recovery?**

*Transit Authority of Central Kentucky Vanpool has the highest farebox recovery at 2.38.*

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS |> select(Agency, Mode, `Total Fares`, Expenses) |> 
  mutate(farebox_recovery = `Total Fares`/Expenses) |> 
  arrange(desc(farebox_recovery))|>
  slice(1)
```

**Which transit system (agency and mode) has the lowest expenses per UPT?**

*San Francisco Bay Area Rapid Transit District Heavy Rail has the lowest expenses per UPT at 0.396.*

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS |> select(Agency, Mode, UPT, Expenses) |> 
  mutate(Expenses_per_UPT = Expenses/UPT) |> 
  arrange(Expenses_per_UPT)|>
  slice(1)
```

**Which transit system (agency and mode) has the highest total fares per UPT?**

*The highest total fares per UPT belongs to Altoona Metro Transit's Demand Response at 656 per UPT.*

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS |> select(Agency, Mode, `Total Fares`, UPT) |> 
  mutate(Total_Fares_per_UPT = `Total Fares`/UPT) |> 
  arrange(desc(Total_Fares_per_UPT))|>
  slice(1)
```

**Which transit system (agency and mode) has the lowest expenses per VRM?**

*San Francisco Bay Area Rapid Transit District's Heavy Rail at 0.217 per VRM.*

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS |> select(Agency, Mode, Expenses, VRM) |> 
  mutate(Expense_VRM = Expenses/VRM) |> 
  arrange(Expense_VRM)|>
  slice(1)
```

**Which transit system (agency and mode) has the highest total fares per VRM?**

*Chicago Water Taxi (Wendella)'s Ferryboat at 237 total fares per VRM*

```{r echo=FALSE, message=FALSE}
USAGE_AND_FINANCIALS |> select(Agency, Mode, `Total Fares`, VRM) |> 
  mutate(Fares_VRM = `Total Fares`/VRM) |>
  arrange(desc(Fares_VRM)) |>
  slice(1)
```

## Conclusion

In terms of ridership, the MTA New York City Transit takes the win in with the most Vehicle Revenue Miles and the most Unlinked Passenger Trips in 2022. Ridership in the NYC, NJ, Newark area overall is the highest and the transit systems in the area are some of the most utilized public transit systems in the US. Financially, San Francisco's BART Heavy Rail/Subway comes out on top with both the lowest expense per VRM and lowest expense per UPT. Additionally, San Francisco/Oakland, CA also offers the most modes of transportation. When it comes to usage, the MTA is the transit system that shines, covering the most revenue miles with its vehicles and servicing the most passenger trips. When finances are added to the picture, the BART seems to be the most cost effective transit system.
